---
title: "European countries' GDP, euro or not"
author: "Mitsuo Shiota"
date: "2019-4-11"
output: 
  github_document:
    toc: TRUE
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

- [European countries' GDP line chart in pdf](output/GDP-euro-or-not.pdf)
- [How much recovered from the Great Recession boxplot, in euro or not, in pdf](output/GDP-euro-or-not2.pdf)

This is my second attempt to utilize R language for economic analysis after [Yellen's US Labor Market Dashboard](yellen-dashboard).

## Libraries

As usual I attach tidyverse package. I also attach rvest package for web scraping to get the euro entry date of each country. Although I don't attach, I use eurostat package to get GDP data, and countrycode package to convert country codes to names.

```{r libraries, message=FALSE}
library(tidyverse)
library(rvest)

```

## Get GDP data

[Eurostat Database](https://ec.europa.eu/eurostat/data/database) provides a wide variety of data. I dig Data navigation tree down to "Quarterly national accounts", reach "GDP and main components (output, expenditure and income) (namq_10_gdp)", and know the table name is "namq_10_gdp". Click Data Explorer icon, and a new window with a large table appears. Look at the upper part, and click + icon to know the parameters, like:

- unit: CLV10_MNAC:	Chain linked volumes (2010), million units of national currency
- s_adj: SCA: Seasonally and calendar adjusted data
- na_item: B1GQ; Gross domestic product at market prices

[Cheat sheet: eurostat R package](https://cran.r-project.org/web/packages/eurostat/vignettes/cheatsheet.html) helps me.

```{r eurostat}
eu_gdp <- eurostat::get_eurostat(id = "namq_10_gdp",
                                 filters = list(
                                   unit = "CLV10_MNAC",
                                   s_adj = "SCA",
                                   na_item = "B1GQ")
                                 )

eu_gdp <- eu_gdp %>% 
  select(time, geo, values)

eu_gdp$geo <- as.character(eu_gdp$geo)

```

## Add country names by looking up codes

I follow the order of ["Tutorial: Country codes and protocol order"](https://ec.europa.eu/eurostat/statistics-explained/index.php/Tutorial:Country_codes_and_protocol_order).

```{r codes_eu28}
country_codes_eu28 <- c("BE", "BG", "CZ", "DK", "DE", "EE", "IE",
                        "EL", "ES", "FR", "HR", "IT", "CY", "LV",
                        "LT", "LU", "HU", "MT", "NL", "AT", "PL", 
                        "PT", "RO", "SI", "SK", "FI", "SE", "UK") 

```

GDP data contain 3 countries outside EU, so I add them to codes.

```{r codes}
country_codes <- c(country_codes_eu28, "NO", "CH", "RS")

```

I create the lookup table of country code and name by utilizing countrycode package.

```{r lookup_tbl}
lookup_tbl <- tibble(
  geo = country_codes,
  name = countrycode::countrycode(country_codes, 'eurostat', 'country.name')
)

```

Then I add "name" column to GDP data by matching code and name, and drop organizations, like EU-28, which don't match country codes.

```{r add_name}
eu_gdp <- eu_gdp %>% 
  left_join(lookup_tbl, by = "geo") %>% 
  drop_na(name)

```

## Which country introduced euro, and when

Now I would like to know when countries switched from national currencies to euro. I use rvest package for web scraping. I have found [wiki page "Euro"](https://en.wikipedia.org/wiki/Euro), and managed to get when euro area countries fixed their currenicies to the euro.

```{r wiki_scrape}
wiki <- read_html("https://en.wikipedia.org/wiki/Euro")

euro_entry <- wiki %>% 
  html_nodes("table") %>% 
  .[[5]] %>% 
  html_nodes("td") %>% 
  html_text() %>% 
  str_sub(1, -2)

euro_entry_tbl <- 
  tibble(
    name = euro_entry[seq(1, 132, 6)],
    date_fixed = euro_entry[seq(1, 132, 6) + 4]
  )

euro_entry_tbl$date_fixed <- as.Date(euro_entry_tbl$date_fixed)

```

I add "date_fixed" column to GDP data, compare it with "time" column, and get "euro" column to show whether each time (row) is in euro ("Y") or not ("N").

```{r add_euro}
eu_gdp <- eu_gdp %>% 
  left_join(euro_entry_tbl, by = "name") %>% 
  mutate(euro = if_else(time >= date_fixed, "Y", "N")) %>% 
  replace_na(list(euro = "N"))

eu_gdp$name <- factor(eu_gdp$name, levels = lookup_tbl$name)

eu_gdp$euro <- factor(eu_gdp$euro)

```

## Plot

I plot quarterly real GDP from 2000 by setting 1Q 2007 (just before the Great Recession) = 100. 

I set the common y-axis range so that I can see the movements of most countries easily. But, as a result, some countries went out of the plot. If you don't like it, set YLIM as you like.

Line is colored differently depending on whether it is in euro or not.

```{r plot, fig.width=10, fig.height=8}
# set the parameters to plot
START = "2000-01-01"
YLIM = c(75, 125)

# index "2007-01-01" = 100
eu_gdp <- eu_gdp %>% 
  group_by(name) %>% 
  mutate(index = values / values[which(time == "2007-01-01")] * 100) %>% 
  ungroup()

# plot
eu_gdp %>% 
  filter(time >= START) %>% 
  ggplot(aes(x = time, y = index)) + 
  geom_hline(yintercept = 100, color = "white", size = 2) +
  geom_line(aes(color = euro), size = 1) +
  facet_wrap(~ name) +
  coord_cartesian(ylim = YLIM) +
  labs(
    title = "Europe; Real GDP, 1Q 2007=100",
    x = "",
    y = "")

```

```{r plot_save}
ggsave(filename = "output/GDP-euro-or-not.pdf",
       width = 10, height = 8, units = "in", dpi = 300)

```

## Euro or not

Can I say something by looking at the plot?

GDP does not reflect employment, if foreigners just borrow the land to produce goods they bring home. GNI would be better for [Leprechaun Economics](https://en.wikipedia.org/wiki/Leprechaun_economics).

Euro area countries are so advanced that they have already reaped low hanging fruits of productivity, and that their demography is not favorable for growth, so their growth path is not steep. On the contrary, some European non-euro counties are still in the developing stage, so their growth path is steep. If I compare how much GDP recovered from the Great Recession between euro and non-euro countries like below, I am not fair.

```{r latest GDP_boxplot, fig.width=3, fig.height=4}
latest_gdp <- eu_gdp %>% 
  group_by(name) %>% 
  filter(time == max(time)) %>% 
  ungroup() %>% 
  mutate(euro = if_else(date_fixed <= "2007-01-01", "Y", "N")) %>% 
  replace_na(list(euro = "N"))

latest_gdp %>% 
  ggplot(aes(x = euro, y = index)) +
  geom_hline(yintercept = 100, color = "white", size = 2) +
  geom_boxplot() +
  labs(
    title = "How much recovered?",
    x = "in euro as of 1Q 2007",
    y = "Latest Real GDP, 1Q 2007=100")

```

```{r boxplot_save}
ggsave(filename = "output/GDP-euro-or-not2.pdf",
       width = 3, height = 4, units = "in", dpi = 300)

```

Who are the winners, and who are the losers?

```{r latest_gdp}
latest_gdp$euro <- factor(latest_gdp$euro)
latest_gdp %>% 
  select(time, name, euro, index) %>% 
  arrange(desc(index))

latest_gdp %>% 
  select(time, name, euro, index) %>% 
  arrange(index)

```

I feel sympathy with Greece and Italy.

EOL
